%The mandatory template presets
\PassOptionsToPackage{table,xcdraw}{xcolor}
\documentclass{article}
\usepackage{fullpage}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{parskip}
\usepackage{hyperref}
\usepackage{pdfpages}
\usepackage[table,xcdraw]{xcolor}
\usepackage{pdflscape}
\usepackage{array}
\usepackage[numbers]{natbib}       % or just use plain \cite

%NEW - Use british english.
\usepackage[british]{babel}

%NEW - allow use of N,C,Q,R, etc..
\usepackage{amsfonts}



%NEW - allows for use of theorems, definitions, and so on.
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}
\newtheorem{definition}{Definition}[section]
\newtheorem{comment}{Comment}
\newtheorem{conjecture}{Conjecture}


\title{TLS report}

\author{Yves Greatti, Thomas Greatrix, Smurity Karale}
\date{May 2025}

\begin{document}

\maketitle

Whilst Machine Learning (ML) techniques are used within the considered literature, the vast majority of papers use them only for statistical analysis. This, in turn, is done to identify predictors of TLS and TLS-related complications. The exceptions to this are ,  \cite{montesinos2008tumor}, \cite{mato2004predictive}, and  \cite{xiao2024prediction}. Our analysis begins by exploring \cite{montesinos2008tumor}, 
\cite{xiao2024prediction}, and  \cite{mato2004predictive} individually, and then explores how ML techniques have been used to perform statistical analysis to locate predictors of TLS.

\section{ML for prediction of TLS}

\cite{montesinos2008tumor} Took data from 194 patients between the ages of 18 and 86 (of which 19 also had TLS) with Acute Myelogenous Leukemia (AML) and performed statistical analysis on this data in order to find predictors of Tumor Lysis Syndrome (TLS). The authors performed both univariate and multivariate analysis, which were done via logistic regression, a binary classification model, which works by fitting a logistic curve that splits the data into two user-defined categories, in this case, patients that have TLS and patients that do not have TLS. By looking at how important specific variables are to the model’s predictions, one can infer the likelihood that a variable is a predictor of TLS. In this case, the authors found that uric acid “(UA) (p=.0003), Cr (p=.0025), [lactate dehydrogenase] LDH (p=.0001), [White Blood Cell count] WBC (p =.0058), gender (p =.0064), and [Chronic Myelomonocytic Leukemia] CMML (p=.0292)” were predictors of TLS. When the authors used multivariate analysis, LDH “(p=0.01, OR 3.01, 95\% CI 1.5–6.2) and UA p =0.01, OR 2.00, 95\% CI 1.4–2.8)” were predictors. While these values are statistically significant, the authors did not mention effect size.

There is some debate on how much data is required for logistic regression, a general rule of thumb is given by:

\[\frac{10C}{P}\]

Where C is the number of covariates (potential predictors of TLS in this case) and P is the smallest proportion of negative or positive cases. It is not stated precisely how many predictors were tested at once during multivariate analysis, but as 6 were found, we can assume \(C \geq 6\). Hence:

\[P = \frac{19}{194} \approx 0.1 \]
\[\frac{10C}{P} \approx \frac{10 \times 6}{0.1} = 600\]

The study had 194 datapoints and used less data than might be recommended for their multivariate logistic regression analysis. When using less data than recommended, there is a risk of the results not being representative of the general population; however, in some cases (especially within healthcare), such a situation is not always avoidable. Their univariate logistic regression did not have this issue, as for C = 1, we find that approximately 100 datapoints would be recommendable. As such, there may be precedence for UA, LDH, WBC,  sex, and CMML to be predictors of TLS.

After using statistical analysis to find predictors of TLS, the authors used LDH and UA to make a points-based predictive model of TLS (PPS-TLS score). Points were assigned based on LDH and UA values, with a bias given to UA. The PPS‑TLS score varies between 0 and 6, and is challenging to use it as a binary classifier: if you set a low threshold (e.g., score \(\ge 2\) or 3), you will catch all the TLS cases but you’ll have so many positives that the test doesn’t significantly change the pre‑test odds \((\mathrm{LR}+ \sim1 \text{--}1.7)\). The median of the patients with TLS was 5, with a sensitivity of 0.63 and a specificity of 0.85, showing that the score has modest discriminative ability. 
A more rigorous approach would be to plot the whole ROC curve (and report the AUC), choose the threshold that maximizes Youden’s \(J\) index, and then assess calibration or validate that cut‑off in an independent sample.

Whilst it is easy to interpret a system wherein predictors are assigned points, the accuracy of the model and the fact that it was based on only two predictors found via multivariate analysis on a small dataset means that it is unlikely to be helpful in a clinical setting. The authors also state that the model has not been externally validated. 

\cite{xiao2024prediction} took 772 adult patients, of whom 130 had TLS. TLS patients were divided into laboratory TLS (LTLS) or clinical TLS (CTLS). The authors began by making all their continuous variables into categorical intervals before doing a univariate analysis to find important variables, which was accomplished via a chi-squared test. Afterwards, the statistically significant variables were used to build a stepwise logistic regression model. Using our rule of thumb from earlier:
\[P =\frac{130}{772} \approx 0.17 \]
\[ \frac{10C}{P} \approx \frac{10\times9}{0.17} \approx 529 \]
Hence, the authors likely had enough data to create such a model. From their multivariate analysis, the authors found that WBC, UA, and LDH were predictors of TLS. 
The CTLS scoring system was established based on the regression coefficients of the multivariate analysis and has good performance (AUC. .81 \%, 95\% bias\-corrected CI, 0.77 to 0.84) on the test dataset, although goodness of fit Hosmer-Lemeshow statistic was not significant (\(\tilde{\chi}^2\)=7.6; p = 0.18).At cut\-off levels at 2 and 3 points, the model achieved sensitivities of 95\% and 89\%, and specificities of 67\% and 80\%, respectively, for predicting clinical TLS. Although assigning discrete points to each predictor makes the score intuitive, its overall accuracy is not modest; it may be better deployed within a broader, multifactorial diagnostic framework.

\cite{mato2004predictive} took data from 2,243 patients under 18 with acute lymphoblastic leukemia (ALL), of which 199 had TLS. The authors used this data to train a variety of machine learning models, to predict whether a patient has TLS. It is worth noting that the group used a miss-forest model to fill in missing values within the dataset. However, this is unlikely to affect the overall results from a machine learning perspective.

To start the study, the authors used LASSO regression to find predictors of TLS. LASSO regression is also known as L1 regression. LASSO regression minimizes the magnitude of coefficients in a Machine Learning model by punishing the model by adding the absolute value of the coefficients multiplied by some coefficient to the loss of the model. At a high level, this incentivizes the model to send the coefficients of unimportant variables to 0. As a result, by performing the LASSO regression, it is possible to tell which variables are insignificant as they will have been set to 0 during the LASSO regression. It is important to note that if two variables are highly correlated (e.g., age and weight in children), LASSO regression may shrink one to zero, even if both contribute value to the model.
Additionally, selecting variables set to zero depends on the training data, making it challenging to interpret why LASSO prioritizes certain features over others. The authors of the paper minimized these disadvantages by using 10-fold cross-validation. However, there are still risks that should be considered when constructing models. The authors found that the LASSO regression selected 'FAB type, WBC, phosphorus, calcium, potassium, UA, AST, blood glucose, occurrence of infection, AKI, cardiac arrhythmia, and type of steroid used in the initial two induction chemotherapy' as the 12 most important variables. WBC was also found as a predictor of TLS in both (Tony H. Truong, 2007, \cite{truong2007features} and \cite{mato2004predictive}. In addition, Anthony R. Mato (2004) found UA to be a predictor of TLS, although this was in a study performed on adults rather than children, with a different form of cancer.

After using LASSO regression to find the 12 most important variables, the researchers created four different models to predict TLS from these variables: CatBoost, logistic regression, random forest, and a Support Vector Machine (SVM). Exploring multiple models is good practice, as it is often difficult to tell how a specific model structure will perform on a dataset without testing it. While it is commendable that they tested multiple hyperparameter configurations for each model to ensure fairness, the authors do not appear to have evaluated the finalized models on various training datasets, ensuring no model was unfairly disadvantaged. 

 It is essential to recognize that model performance can be heavily influenced by the training data used and the initial conditions of a given run, which are typically random. Without repeated testing, there is a risk that model performance only varies due to data differences. This is especially relevant since the models analyzed by the authors performed similarly, so changes to the training data and running repeated tests may have changed the results.
It is also important to note that after identifying the twelve predictors of TLS, the authors did not investigate their biological and clinical significance. Without a strong biological argument, the paper loses some potential impact.

Ultimately, the paper shows mostly good practice within AI and uses appropriate techniques.  Although CatBoost outperformed the other models in the paper, this may not always hold true in practice. Additionally, while the authors suggest that CatBoost "can be incorporated into a clinical decision support system," its accuracy was only 75.7\%. While CatBoost predictions are theoretically interpretable, understanding the reasoning behind its predictions in practice can be challenging, which may limit its practical applicability.

\section{ML for statistical analysis of TLS factors}

%%%%%
%%%%%
%%%%%
%%%%%

Within the considered literature, the vast majority of authors are focused on finding predictors of TLS and TLS-related complications. The methods used for this are typically well-established statistical techniques, with a heavy emphasis on logistic regression and, to a lesser extent, COX regression. Below, we have provided a table that provides information about each of the papers. Note that \cite{mato2004predictive}, 
\cite{montesinos2008tumor}, and  \cite{xiao2024prediction} were discussed earlier, and are thus excluded from the table.

\begin{landscape}
\begin{table}[]
\hspace*{-3.5cm}\begin{tabular}{>{\raggedright\arraybackslash}p{1cm}|l|l|l|l|l|l|}
\hline
\cellcolor[HTML]{FFFFFF}{\color[HTML]{333333} Name} & 
Model(s) used & 
\begin{tabular}[c]{@{}l@{}}Number of\\ patients\\ in study.
\end{tabular} & 
\begin{tabular}[c]{@{}l@{}}Number of TLS\\ patients in study\end{tabular} & Finding & Notes                                                               \\
\hline
\multicolumn{1}{|l|}{\begin{tabular}[c]{@{}l@{}}Features at presentation predict children\\ with acute lymphoblastic leukemia at low\\ risk for tumor lysis syndrome.\\ (Tony H. Truong, 2007)\end{tabular}}           
& \begin{tabular}[c]{@{}l@{}}Logistic regression\\ (both univariate and\\ multivariate)\end{tabular}
& 74
& 328
& \begin{tabular}[c]{@{}l@{}}“sex, age, WBC, mediastinal mass,\\ hepatomegaly, splenomegaly, and\\ T-cell immunophenotype” were\\ all predictors of TLS.\end{tabular}
& N/A

\\
\hline
\multicolumn{1}{|l|}{\begin{tabular}[c]{@{}l@{}}Serum phosphate level and its kinetic as\\ an early marker of acute kidney injury\\ in tumor lysis syndrome (Marie Lemerle,\\ 2022)\end{tabular}}
& COX regression
& 120
& 120
& \begin{tabular}[c]{@{}l@{}}“increases in serum phosphate and\\ LDH appear to be early and reliable\\ biomarkers of AKI in tumour lysis\\ syndrome”\end{tabular}         
& \begin{tabular}[c]{@{}l@{}}Studied acute kidney injury\\ predominantly, instead of the\\ factors that cause TLS.\end{tabular}

\\
\hline
\multicolumn{1}{|l|}{\begin{tabular}[c]{@{}l@{}}(Fausto A. Rios-Olais, 2024) Tumor Lysis\\ Syndrome Is Associated with Worse\\ Outcomes in Adult Patients with Acute\\ Lymphoblastic Leukemia. (Fausto A. Rios\\ -Olais, 2024)\end{tabular}}
& \begin{tabular}[c]{@{}l@{}}Logistic regression\\ (Multivariate), COX\\ regression.\end{tabular}
& 138
& 42                                   & \begin{tabular}[c]{@{}l@{}}UA, LDH, and Male sex were\\ significant predictors of clinical\\ TLS. LDH and WBC were\\ significant predictors of TLS.\end{tabular}
& N/A

\\
\hline
\multicolumn{1}{|l|}{\begin{tabular}[c]{@{}l@{}}Uric Acid and the Prediction Models of\\ Tumor Lysis Syndrome in AML (A. \\Ahsan Ejaz, 2015)\end{tabular}}
& \begin{tabular}[c]{@{}l@{}}Logistic regression\\ (Multinomial)\end{tabular}
& 183
& \begin{tabular}[c]{@{}l@{}}48 LTLS and\\ 10 CTLS\end{tabular}
& \begin{tabular}[c]{@{}l@{}}WBC is a better predictor of TLS\\ than other predictors.\end{tabular}
& N/A                                                                        
\\
\hline
\multicolumn{1}{|l|}{\begin{tabular}[c]{@{}l@{}}Predictors for Severe Tumor\\ Lysis Syndrome (Scott Wirth, 2012)\end{tabular}}
& Logistic regression
& 1327
& N/A                                  & \begin{tabular}[c]{@{}l@{}}Sex, age, cancer type, risk category,\\  Serum Creatinine levels, Blood\\ Urea Nitrogen levels, Magnesium\\ levels, Phosphorus levels, all\\ predicted severe outcomes.\end{tabular}
& \begin{tabular}[c]{@{}l@{}}Studied the likelihood of a\\ severe outcome based on\\ risk category and other\\ standard predictors.\end{tabular}

\\
\hline
\multicolumn{1}{|l|}{\begin{tabular}[c]{@{}l@{}}Risk factors and development of a\\ predictive score model for tumor lysis\\ syndrome in childhood leukemia: a\\ 10-year experience from a single tertiary\\ hospital in Thailand (Pharsai Prasertsan,\\ 2024)\end{tabular}} & \begin{tabular}[c]{@{}l@{}}Logistic regression\\ (univariate,\\ multivariate)\end{tabular}
& 252
& \begin{tabular}[c]{@{}l@{}}51 with TLS, of\\ which 24 had\\ CTLS\end{tabular}
& \begin{tabular}[c]{@{}l@{}}Age, BMI, Mediastinal mass,\\ WBC, LDH, GFR, AST\\ were all identified as risk factors\\ for TLS.\end{tabular}        & N/A                                  

\\
\hline
\multicolumn{1}{|l|}{\begin{tabular}[c]{@{}l@{}}Risk factors for tumor lysis syndrome in\\patients with chronic lymphocytic\\leukemia treated with the cyclin-\\dependent kinase inhibitor,\\ flavopiridol (K A Blum, 2011)\end{tabular}}
& \begin{tabular}[c]{@{}l@{}}Logistic regression\\(univariate,\\multivariate)\end{tabular}
& 116
& \begin{tabular}[c]{@{}l@{}}53\end{tabular}
& \begin{tabular}[c]{@{}l@{}}The authors found 5 statistically\\significant risk factors for TLS:\\WBC, gender, Bulky\\lymphadenopathy, \(\beta\)2-microglobulin,\\ Albumin.\end{tabular} & N/A       

\\
\hline
\multicolumn{1}{|l|}{\begin{tabular}[c]{@{}l@{}}In‐Hospital Outcomes of Tumor Lysis\\Syndrome: A Population‐Based Study\\Using the National Inpatient Sample\\(Urshila Durani, 2017)\end{tabular}}
& \begin{tabular}[c]{@{}l@{}}Logistic regression,\\ linear regression.\end{tabular}
& 28,370
& \begin{tabular}[c]{@{}l@{}}28,370\end{tabular}
& \begin{tabular}[c]{@{}l@{}}“age, weighted Elixhauser comorbidity\\score, insurance, teaching hospital\\ versus non teaching hospital, and\\cancer type were predictors of in‐\\hospital mortality, whereas sex, race, \\facility, and income were not”\end{tabular} & N/A       

\\
\hline
\end{tabular}

\end{table}

\begin{table}[]
\hspace*{-3.5cm}\begin{tabular}{>{\raggedright\arraybackslash}p{1cm}|l|l|l|l|l|l|}
\hline
\cellcolor[HTML]{FFFFFF}{\color[HTML]{333333} Name} & 
Model(s) used & 
\begin{tabular}[c]{@{}l@{}}Number of\\ patients\\ in study.
\end{tabular} & 
\begin{tabular}[c]{@{}l@{}}Number of TLS\\ patients in study\end{tabular} & Finding & Notes                                                               \\
\hline
\multicolumn{1}{|l|}{\begin{tabular}[c]{@{}l@{}}Predictors of in-Hospital Mortality in\\ Tumor Lysis Syndrome \end{tabular}}           
& \begin{tabular}[c]{@{}l@{}}Logistic regression\\ (univariate and\\ multivariate)\end{tabular}
& 997
& 997
& \begin{tabular}[c]{@{}l@{}}“Independent predictors of in-hospital\\ mortality were cardiac dysrhythmias\\ and sepsis."\end{tabular}
& N/A

\\
\hline
\multicolumn{1}{|l|}{\begin{tabular}[c]{@{}l@{}}Identification of Children with Acute\\ Lymphoblastic Leukemia at Low Risk\\ for Tumor Lysis Syndrome (Bahoush \\GR, 2015)
 \end{tabular}}           
& \begin{tabular}[c]{@{}l@{}}Logistic regression\\ (univariate and\\ multivariate)\end{tabular}
& 160
& 41
& \begin{tabular}[c]{@{}l@{}}“CNS and renal involvement and\\ T-cell immunophenotype are among\\ the strongest predictors of TLS."\end{tabular}
& N/A

\\
\hline
\multicolumn{1}{|l|}{\begin{tabular}[c]{@{}l@{}}Risk-based management strategy and\\ outcomes of tumor lysis syndrome in\\ children with leukemia/lymphoma:\\ Analysis from a resource-limited\\ setting
(Gopakumar, 2018) \end{tabular}}
& \begin{tabular}[c]{@{}l@{}}Logistic regression\\ (Multivariate)\end{tabular}
& 224
& 41
& \begin{tabular}[c]{@{}l@{}}“Maintaining hydration and\\ establishing adequate urine output\\ prior to chemotherapy along with the\\ judicious use of rasburicase helps in\\ managing TLS in an economically\\ viable manner in a resource-limited \\setting.”\end{tabular}
& N/A

\\
\hline
\multicolumn{1}{|l|}{\begin{tabular}[c]{@{}l@{}}Plasma uric acid response to rasburicase:\\ early marker for acute kidney injury in\\ tumor lysis syndrome? (Canet, 2014)
 \end{tabular}}
& \begin{tabular}[c]{@{}l@{}}Logistic regression\end{tabular}
& 60
& \(\approx\) 40 (unspecified)
& \begin{tabular}[c]{@{}l@{}}“Baseline plasma uric acid was higher\\ in the AKI group.” and "Conceivably,\\ a smaller response to rasburicase may\\ indicate subclinical AKI at presentation,\\ with a decrease in uric acid clearance."\end{tabular}
& N/A

\\
\hline
\end{tabular}

\end{table}
\end{landscape}

\section{Discussion}

The most common issue among the papers considered is the small datasets. Many of the papers considered do not have enough data to make reliable claims using logistic regression or more traditional statistical testing, and as such, they risk having results that cannot be reproduced. It is worth noting that this problem is widespread within the medical field due to the ethical and logistical issues with data collection.

The most common approach in the papers involved using logistic regression, possibly with some pre-processing, to find potential predictors of TLS. Such an approach is well-founded from a mathematical standpoint. However, many assumptions are made when using logistic regression. First, logistic regression is limited in how complex the relationships between variables can be, so more complex relationships between variables may be missed. Logistic regression is also sensitive to outliers, and given the small sample sizes used and dealing with complex real-world datasets, it is not unreasonable to expect outliers to be present in the data, nor is it unreasonable to assume that such outliers might be difficult to detect reliably. The usage of logistic regression for univariate and multivariate statistical testing, which is common in the papers considered, is not always a good approach (especially in the univariate case), as the models created are typically more complex to interpret and reproduce, and often are less robust and reliable for statistical testing than more traditional statistical tests. The decision to develop models capable of making predictions about TLS outcomes and then not evaluating them on their ability to do so is also questionable - especially when the primary purpose of these models is to do so.

Some papers also explored the usage of COX regression to find how variables may affect the time for important events to occur. COX regression, similarly to logistic regression, is also sensitive to outliers. COX regression also assumes that the effect a variable has is constant over time. As a toy example, COX regression may be able to tell you that people who smoke are more likely to develop lung cancer. However, COX regression may also tell you that someone who has been smoking for 5 years is just as likely to develop cancer as someone who started yesterday. Whilst COX regression is well-established, authors typically do not acknowledge or account for these drawbacks in their analysis, which could make results unreliable and miss potentially useful details.

In some of the papers considered, researchers also explored models that predicted the likelihood of TLS occurring based on provided variables. These approaches were typically point-based systems, wherein variables were assigned “points” based on their value. The points would then be added up, and based on the total, the risk of a specific outcome could be calculated. This approach is inherently well-suited for clinical environments, as it is extremely easy to interpret the result and see the most significant contributing factors to the result. Unfortunately, such models were generally not accurate enough to be clinically useful as a standalone tool, and it is recommended that the models be used as part of a larger diagnosis procedure instead. \cite{mato2004predictive}  compared a range of well-established, but far less interpretable, AI techniques and found that CATboost had the best performance for predicting TLS. However, this approach was still unlikely to be accurate enough to be clinically useful as a standalone tool.

While the methods used in the papers considered are in theory explainable, it is worth noting that it may still be difficult in practice to tell why a model made a specific connection between variables - especially if a large amount of variables are present. This being said, the final result of the models (i.e., tumour size may be a predictor of TLS) is understandable and in a worst-case scenario can be verified experimentally after the analysis has taken place. There is also an argument to be made that any analysis that includes large numbers of variables will inherently be challenging to understand. In practice, these models are complex, can find surprising links, and sometimes find false relationships. Performing reality checks on the results, as well as potential follow-up studies and analysis, means that the results from these models can be valuable for advancing TLS research.

It is worth mentioning that it is difficult to tell how well the models created will perform in real-world clinical settings. The accuracy of the models is often not checked, and the data is not released publicly, making it hard to validate what has been done. The papers also do not create more than one model, or use multiple randomised test/train sets, which is standard within the Machine Learning field to test reliability. The small datasets used mean that models may not perform well for the general population, and the models have significant risks of unexpected behaviour in edge cases, extreme cases, and unusual cases.

In the future, the field may want to consider how more sophisticated points-based TLS prediction models could be developed, which may be particularly well suited to clinical settings due to being highly interpretable for both specialists and non-specialists. If this turns out to be unfeasible, the usage of ensemble learning (the usage of multiple different models to make predictions) and transfer learning (taking a big, generalist model and training it to work well in your specific domain) have both been shown to perform well in the sort of low-data situations that appear to be common within the field.


\newpage
\bibliographystyle{unsrtnat}      
\bibliography{bibliography}

\end{document}

